# Practical 1: A control chart problem

## part a - summary of waiting times

```{r, echo = FALSE, message = FALSE, include=FALSE}
source(here::here("scripts/setup.R"))
```


```{r, warning=FALSE, message=FALSE}
waiting <- read.csv(here::here("data/waiting.csv"))
```

The *Graph* displays the daily waiting times period that the Data covers. It is difficult to notice any seasonality or cyclic patterns in the extreme waiting times.

```{r}
waiting$Date <- as.Date(waiting$Date)

waiting %>%
  ggplot(aes(Date, Average.Wait.Seconds)) +
  geom_line() +
  geom_point() +
  labs (title = "Average daily waiting time",
        x = "Date",
        y = "Time [seconds]")
```

As shown on the *boxplot* and on the *table*, Monday, Tuesday and Wednesday are the days of the week tht present the highests extreme values. We can also notice that even if Friday has the highest average waiting time, its maximum value is the lowest.

Moreover, the *boxplot* shows that the first 4 days of the week there are outliers, which are the part values that we analyse when working with extreme values.

```{r}
waiting %>%
  mutate(Weekday=factor(Weekday, levels= c("Monday", 
    "Tuesday", "Wednesday", "Thursday", "Friday"))) %>%
  group_by(Weekday) %>%
  summarize(Min = min(Average.Wait.Seconds),
             Q1 = quantile(Average.Wait.Seconds, .25),
             Avg = mean(Average.Wait.Seconds), 
             Q3 = quantile(Average.Wait.Seconds, .75),
             Max = max(Average.Wait.Seconds)) %>%
  kable_maker(caption="Data summary")
```

```{r}
waiting %>%
  mutate(Weekday=factor(Weekday, levels= c("Monday", 
    "Tuesday", "Wednesday", "Thursday", "Friday"))) %>%
  ggplot(aes(Average.Wait.Seconds, Weekday)) +
  geom_boxplot() +
  coord_flip() +
  labs (title = "Distribution of waiting times per day",
        y = "Day",
        x = "Time [seconds]")
```

## part b - normal approximation

```{r}
n <- 250

p <- (1-(1/n))
```

As the one year period corresponds to  $$n = 250$$, we can deduce the upper control limit, which corresponds to $$p = 1-1/n$$ and equals to `r p`. 
After scaling this value, we can retrieve the associated quantile or the upper control limit.
```{r}
mean <- mean(waiting$Average.Wait.Seconds)
sd <- sd(waiting$Average.Wait.Seconds)

upper_limit <- qnorm(p, mean, sd)
upper_limit %>% 
  kable_maker(caption = "Upper Limit")
```

The upper control limit corresponds to `r upper_limit`.

The *Graph* suggest that the normal distribution is not appropriate to predict extreme values, the extreme observations tend to diverge strongly from their theorical values under the assumption of a normal distribution. This suggests that the tails are heavy.
```{r}
qqnorm(scale(waiting$Average.Wait.Seconds))
qqline(scale(waiting$Average.Wait.Seconds))
hist(waiting$Average.Wait.Seconds, main = "Histogram", xlab = "Time [seconds]") 
```

```{r}
shapiro <- shapiro.test(waiting$Average.Wait.Seconds)
```

Running the Shapiro-Wilk test shows that the hypothesis of normality is rejected at level `r shapiro$p.value`

## part c - analzying long waiting times 
### Block Maxima    

The Block Maxima approach divides the dataset into blocks with the same number of observations. Then, for each block we select the observation with the highest value. Looking at the dataset, we can divide it by months or on a weekly basis :

```{r}
plot_all_obs <-waiting %>% 
  ggplot(aes(x = Date, y = Average.Wait.Seconds)) +
  geom_point()

tmp <- data.frame(x=rep(seq(as.Date(0, origin="2017-01-01"),
                            length=36, by="1 month"), 2),
                  y=rnorm(72),
                  category=gl(2,36))

plot_all_obs + geom_vline(xintercept=as.numeric(tmp$x),
                linetype=4, colour="red") + labs(x = "Date", y = "Time [seconds]", title = "Data division using the Block Maxima", subtitle = "Division on a monthly basis")
```


```{r}
# per week
tmp <- data.frame(x=rep(seq(as.Date(0, origin="2017-01-01"),
                            length=96, by="1 week"), 2))

plot_all_obs + geom_vline(xintercept=as.numeric(tmp$x),
                linetype=4, colour="red") + labs(x = "Date", y = "Time [seconds]", title = "Data division using the Block Maxima", subtitle = "Division on a weekly basis")
```


Looking at the data aggregation per month and week, we can say that the first seems to be better. However in order to confirm this hypothesis, when modelling we are going to analyse both cases and keep the model that provides the best results.
For the models, we are going to fit a GEV function on the maxima of each block, represented by the red dots in the following plots.

```{r}
maxima_monthly <- waiting %>% 
  group_by(year(ymd(Date)), month(ymd(Date))) %>% 
  filter(Average.Wait.Seconds == max(Average.Wait.Seconds))

maxima_weekly <- waiting %>% 
  group_by(year(ymd(Date)), week(ymd(Date))) %>% 
  filter(Average.Wait.Seconds == max(Average.Wait.Seconds))
```


```{r}
waiting %>%
  ggplot(aes(Date, Average.Wait.Seconds)) +
  geom_line() +
  geom_point(aes(Date, Average.Wait.Seconds), data = maxima_monthly, color = "red") +
  labs (x = "Date", 
        y = "Time [seconds]", 
        title = "For each block, the red point represents its maxima", 
        subtitle = "Division by month")
```

```{r}
waiting %>%
  ggplot(aes(Date, Average.Wait.Seconds)) +
  geom_line() +
  geom_point(aes(Date, Average.Wait.Seconds), data = maxima_weekly, color = "red") +
  labs (x = "Date", 
        y = "Time [seconds]", 
        title = "For each block, the red point represents its maxima", 
        subtitle = "Division by week")
```

### Peaks-over-threshold approach

The Peaks-over-threshold approach defines the extreme values as the observations above a certain threshold u. In order to find this threshold, we can use an mrlplot and the tcplot. 

```{r,message = FALSE,warning=FALSE}
mrlplot(waiting$Average.Wait.Seconds,umin = min(waiting$Average.Wait.Seconds), umax = max(waiting$Average.Wait.Seconds))

tlim <- c(81,1182)
tcplot(waiting$Average.Wait.Seconds, tlim = c(81,1100), std.err = FALSE)

```
Looking at the plot, we believe that the best threshold is at a value of 800. However, we are going to see if this holds true by analyzing the results in case we use a threshold of 600.

```{r,message = FALSE,warning=FALSE}
u1 <- 600

u2 <- 800

waiting$col1 <- cut(waiting$Average.Wait.Seconds,
               breaks = c(-Inf, u1, Inf),
               labels = c("<=u", ">u")) 

waiting$col2 <- cut(waiting$Average.Wait.Seconds,
               breaks = c(-Inf, u2, Inf),
               labels = c("<=u", ">u"))
```


```{r,message = FALSE,warning=FALSE}
waiting %>%
  ggplot() +
  geom_line(aes(Date, Average.Wait.Seconds)) +
  geom_point(aes(Date, Average.Wait.Seconds, color = col1)) +
  scale_colour_manual(values = c('black', 'red')) +
  geom_hline(yintercept = u1,
             linetype = 2,
             colour = "red") +
  guides(color = FALSE) +
  labs (x = "Date", 
        y = "Time [seconds]", 
        subtitle = "In red, the observation above the threshold u = 600",
        title = "Average daily waiting time")
```

```{r}
waiting %>%
  ggplot() +
  geom_line(aes(Date, Average.Wait.Seconds)) +
  geom_point(aes(Date, Average.Wait.Seconds, color = col2)) +
  scale_colour_manual(values = c('black', 'red')) +
  geom_hline(yintercept = u2,
             linetype = 2,
             colour = "red") +
  guides(color = FALSE) +
    labs (x = "Date", 
        y = "Time [seconds]", 
        subtitle = "In red, the observation above the threshold u = 800",
        title = "Average daily waiting time")
```

We are going to fit a GPD on the difference between the observations highleted in the red dots and the thresholds. Once, using a threshold of 800 and then, of 600.

## part d - modelling the extremes 
### Block Maxima

To see what kind of distribution to use, we have to estimate $\xi, $\sigma, and $\mu for the maxima. First, we will fit a model on the monthly maxima, then on the weekly ones.

Monthly :
```{r, warning = FALSE, message = FALSE}
# fitting the GEV to the maxima in order to have the location, scale and shape parameterss
fit_gev_maxima_monthly <- fgev(maxima_monthly$Average.Wait.Seconds)
fit_gev_maxima_monthly
shape <- fit_gev_maxima_monthly$estimate[3]
```

The estimates of the GEV show that the latter is a Weibull: indeed, the shape parameter is negative (`r shape`).


```{r, fig.asp= 1.2}
par(mfrow = c(2,2))
plot(fit_gev_maxima_monthly)
```

Looking at the plot, we remark that if on the one hand the probability plot is not well fitted, on the other hand the observations almost perfectly fit the straight line in the qq-plot. Moreover, the Return Level Plot shows a 
function that is concave: this is a consequence of the negative shape parameter. 

We are now going to make the same analysis but this time using the weekly maxima.

Weekly:

```{r, warning = FALSE, message = FALSE}
fit_gev_maxima_weekly <- fgev(maxima_weekly$Average.Wait.Seconds)
fit_gev_maxima_weekly
```

Once again, we have a shape parameter that is negative, meaning that we are still in a Weibull. 

```{r,fig.asp= 1.2}
par(mfrow = c(2,2))
plot(fit_gev_maxima_weekly)
```

Comparing the results between monthly blocks and weekly ones, we can clearly see that the results are better in the latter. Indeed, points are almost perfectly fitted for both the probability and quantile plot, which was not the case when we fitted the GEV to the monthly maxima. Thus, we can conclude that this for the block maxima approach the maxima from weekly blocks should be used.

### POT

We are now going to fit a model using the Peak-over-threshold approach. The function we use for this model is fpot, which fits the difference between the observations and the threshold to a GPD. 
As we saw previously, two thresholds are likely to be interesting for the modeling: the first one is a threhsold u1 of 800, and the second is at 600. Like we did in the previous analysis with the block maxima, we are going to analyse both and select the best model.


```{r}
# fits the w = x - u to a gpd
fit_pot_u1 <-fpot(waiting$Average.Wait.Seconds, threshold = u1, model = c("gpd"))
fit_pot_u2 <-fpot(waiting$Average.Wait.Seconds, threshold = u2, model = c("gpd"))
```

```{r, fig.asp=1.2}
par(mfrow=c(2,2))
plot(fit_pot_u1)
```
The plots above are the results of the analysis using a threhsold of `r u1`. We can see that the output are good, meaning that the model is well fitted to our data. Moreover, we can observe that once again the Return Level Plot has a function that is concave, suggesting a negative shape parameter for the model. 

We are now going to do the same analysis using a threshold of `r u2`

```{r,fig.asp=1.2}
par(mfrow=c(2,2))
plot(fit_pot_u2)
```

Naturally, having a higher threshold, the number of observations is lower. Looking at the results, we can conclude that we should select the threshold `r u1` since it allows having a better pp-plot and qq-plot.

## part e - upper control limit 

Since it is difficult to distingish which is the best model between the Block Maxima (weekly) and the Peaks-over-thresold (with u = 600), we are going to compute confidence line at the one year return level for both approaches.

### Block Maxima

```{r}
# return level block maxima:
return_level_maxima <- fit_gev_maxima_weekly$estimate[1] + (fit_gev_maxima_weekly$estimate[2]/fit_gev_maxima_weekly$estimate[3])*(((-log(1-1/50))^-fit_gev_maxima_weekly$estimate[3])-1)

VaR_BM <- return_level_maxima[[1]]
VaR_BM
```

Using the Block Maxima approach `r VaR_BM` is the average waiting time that is expected to be exceeded once every 250 open days (one year).

```{r}
waiting %>%
  group_by(year(Date)) %>%
  filter(Average.Wait.Seconds > VaR_BM) %>%
  ungroup() %>%
  dplyr::select(Date, Average.Wait.Seconds)
```

Using the historical data, we notice that this thresold as been exceeded twice in two years.

### POT

```{r}
alpha <- p

n.survivors <- waiting %>%
  filter(Average.Wait.Seconds > u1) %>%
  summarize(n = n()) %>%
  pull(n)

p.survivors <- n.survivors/length(waiting$Average.Wait.Seconds)

VaR_POT <- u1 + fit_pot_u1$estimate[1]/fit_pot_u1$estimate[2]*(((1-alpha)/p.survivors)^-fit_pot_u1$estimate[2] -1)
```

Using the POT approach, we conclude that `r VaR_POT` is the average waiting time that is expected to be exceeded once every 250 open days (one year), or `r 1-alpha`% of the time.

To conclude, we prefer to adopt a conservative approach and retain the confidence line with the lowest value, that is `r VaR_BM`, obtained with the Block Maxima approach.



