---
title: "report"
output: html_document
---

```{r, echo = FALSE, message = FALSE, include=FALSE}
source(here::here("scripts/setup.R"))
```

# a)

```{r}
sales_data <-
  read.delim(here::here("data/sales_1.txt"), col.names = "Sales") %>% mutate(Day = row_number())
```

```{r}
stockout <- which(diff(sales_data$Sales) == -sales_data$Sales & sales_data$Sales !=0)

ggplot(data = sales_data, aes(x = Day, y = Sales)) +
  geom_line()+
  geom_point()+
  geom_vline(xintercept =stockout, color= "red") +
  labs(title = "Sales unit by day")+
  theme(plot.title = element_text(hjust = 0.5))
```
The red line shows when the observed number of sales goes to zero, essentially the extreme of the sales, there are also other instances where it nears to zero however we ignore them for simplicity sake. We run of stock in these cases.

# b)
Newsvendor model.

```{r}
# Manual approach
newsvendor_mod <- function(price,cost,unit_salvage){
  salvage <- unit_salvage * cost
  underage <- price - cost
  overage <- cost - salvage
  fractile <- underage / (overage + underage)
  return(fractile)
}

cf_news <- newsvendor_mod(10,1,0.1)

cf_news
```

Alternatively, we could have used the [Scperf](https://cran.r-project.org/web/packages/SCperf/SCperf.pdf) library and the `Newsboy` function.

```{r}
# Using a library
library(SCperf)
mean_s <- mean(sales_data$Sales)
sd_s <- sd(sales_data$Sales)
Newsboy(m= mean_s,sd= sd_s,p=10,c=1,s=0.1)
```

# c)
Value-at-Risk (VaR) also known as chance constraint is used in the newsvendor model as a constraint, limiting the probability of particular event (in this case stock-out) happening.

VaR answers the question of how much one can lose with
x% probability over a given time horizon.

In this case, we have much more underage cost than overage giving us a ratio of `r cf_news`.

Expected shortfall or conditional or sometimes called conditional value-at-risk , is about 'how bad can things get?'

TO BE CONTINUED

# d)


Poisson model.

using maximum likelihood estimation for μ.

Mean excess plot.

`Number of exceedences / n`

```{r}
# Manual approach
newsvendor_mod <- function(price,cost,unit_salvage){
  salvage <- unit_salvage * cost
  underage <- price - cost
  overage <- cost - salvage
  fractile <- underage / (overage + underage)
  return(fractile)
}

cf_news <- newsvendor_mod(10,1,0.1)

cf_news
```


```{r}
library (MASS)
sales_poisson <- MASS::fitdistr(sales_data$Sales, "Poisson")

qpois(cf_news,sales_poisson$estimate)


-((log(1-cf_news))/sales_poisson$estimate)
  
# ppois(quant,sales_poisson$estimate, lower.tail= F)

# 1- probability of a certain demand


# qpois(cf_news,sales_poisson$estimate, lower.tail = F)

# Ratio
sales_poisson

```


# e)

POT model.

First we have to determine a value for `u`.

```{r,message = FALSE,warning=FALSE}
u_plot <- function (a_column) {
  min_col <- min(a_column)
  max_col <- max(a_column)
  mrlplot(a_column,
          umin = min_col,
          umax = max_col)
  upper_threshold <- (max_col-(max_col*0.1))
  tcplot(a_column,
         tlim = c(min_col, upper_threshold),
         std.err = FALSE)
}

u_plot(sales_data$Sales)

```

Thresholds of 150 and 170 seem interesting, we can test for both.

```{r}
# different_u <- list(u150=(150, u170= 170)

u150 <- 150
u170 <- 170

# Later have to use a mutate+ across + functions in order no to have code duplicates 
sales_data$col1 <- cut(sales_data$Sales,
               breaks = c(-Inf, u150, Inf),
               labels = c("<=u", ">u")) 

sales_data$col2 <- cut(sales_data$Sales,
               breaks = c(-Inf, u170, Inf),
               labels = c("<=u", ">u"))

cutter <- function(a_column, u_threshold,a_name){
  sales_data$hi <- cut(sales_data$Sales,
               breaks = c(-Inf, u_threshold, Inf),
               labels = c("<=u", ">u"))
}

sales_data %>%
  ggplot() +
  geom_line(aes(Day, Sales)) +
  geom_point(aes(Day, Sales, color = col1)) +
  scale_colour_manual(values = c('black', 'red')) +
  geom_hline(yintercept = u150,
             linetype = 2,
             colour = "red") +
  guides(color = FALSE)

cutter(sales_data$Sales, u170)

sales_data %>%
  ggplot() +
  geom_line(aes(Day, Sales)) +
  geom_point(aes(Day, Sales, color = col2)) +
  scale_colour_manual(values = c('black', 'red')) +
  geom_hline(yintercept = u170,
             linetype = 2,
             colour = "red") +
  guides(color = FALSE)

# The code below is to be removed later
# for (i in seq_along(different_u)){
#   cutter(sales_data$Sales, different_u[[i]])
#   
# }
# cutter(sales_data$Sales, u150)

```

Value at Risk.

```{r}


```

# f)

Back test as seen on the slides. We will go over this in more detail.

```{r}

```



